\bf Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-05-11 doc/paper.Rmd
keywords

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of Statistics, Iowa State University 
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge ...
- affiliation: Department of Statistics and CSAFE, Iowa State University 
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- 3 to 6 keywords
- that do not appear in the title
bibliography: bibliography
abstract: null
---

@@

\gk{The key effect of the optimal locations and thereby the coarseness is seen in the number of failed tests. Irrespective of whether CS1 or CS2 is being used, if the relative optimal locations are at the boundaries we will see an increase in the number of failed tests. A balance is therefore needed in the selection of the coarseness parameter which reduces the boundary effect but also does not remove important (sub-)class characteristics}

@@

>>>>

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of Statistics, Iowa State University 
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge ...
- affiliation: Department of Statistics and CSAFE, Iowa State University 
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- forensic science
- toolmark
- cross-correlation
- Mann-Whitney U statistic
- 
- 
bibliography: bibliography
abstract: null
---

@@

\gk{The key effect of the optimal locations and thereby the coarseness is seen in the number of failed tests. Irrespective of whether CS1 or CS2 is being used, if the relative optimal locations are at the boundaries we will see an increase in the number of failed tests. A balance is therefore needed in the selection of the coarseness parameter which reduces the boundary effect but also does not remove important individual characteristics}

@@

fails <- cvcs %>% group_by(wo, wv, coarse, method, match) %>% summarize(
  fails = 100.0*sum(is.na(p_value))/n()
)

### 2018-05-11 doc/paper.tex
keywords

\newcommand{\hh}[1]{{\textcolor{orange}{#1}}}
\newcommand{\gk}[1]{{\textcolor{green}{#1}}}
\newcommand{\cited}[1]{{\textcolor{red}{#1}}}

\setlength\parindent{0pt}

@@

\section{Introduction and Background}\label{introduction-and-background}

\subsection{Motivation}\label{motivation}

@@

\hh{In the next page and a half it is easy to lose the red line. It might help to include a table with an overview.
The table should include the reference to the paper, the data used, the statistical method and the associated error rates.}
Various toolmarks have been studied in the literature:
\citet{manytoolmarks1} and \citet{chumbley} have been analyzing
screwdriver marks digitized using a profilometer; \citet{manytoolmarks2}
have investigated 3D marks from screwdriver, tongue and groove pliers
captured using a confocal microscope; \citet{afte-chumbley} have been
investigated digitized marks from slip-joint pliers generated by a
surface profilometer.

\hh{We need an additional sentence here to get from the data to the statistical methods ... }
\citet{manytoolmarks2} define a relative distance metric and use it as
similarity measure between two toolmarks. \citet{manytoolmarks1} extract
many small segments in the markings of two toolmarks and compare
similiarity using a maximum pearson correlation coefficient. The
Chumbley scoring method, first introduced by \citet{chumbley}, uses a
similar but more extensive framework based on a Mann-Whitney U test of
the resulting correlation coefficients. This approach is
non-deterministic, because segments are chosen randomly. \citep{hadler}
make the score deterministic for each pair of toolmarks by choosing
segments for comparison systematically. This approach also ensures
independence between segements of striae. In this paper, we are
investigating the applicability of the Chumbley scoring method by
\citet{hadler} to assess striation marks on bullet lands for same-source
identification.

@@

\subsection{Scans for land engraved
areas}\label{scans-for-land-engraved-areas}

@@

\subsection{The Chumbley Score Test}\label{the-chumbley-score-test}

@@

After removing sub-class structure, the Chumbley scores is calculated in
two steps: an optimization step and a validation step. In the
optimization step, the two markings are aligned horizontally such that
within a pre-defined window of length \(w_o\) the correlation between
\(x(t_1)\) and \(y(t_2)\) is maximized: \[
\left(t_1^o, t_2^o\right) = \mathop{\arg \max}\limits_{1 \le t_1 \le T_1, 1 \le t_2 \le T_2} \text{cor} \left(x^{w_o} (t_1), y^{w_o}(t_2) \right)
\] This results in an optimal vertical (in-phase) shift of
\(t_1^o - t_2^o\) for aligning the two markings.
\hh{We will denote the relative optimal locations as $t_1^*$ and $t_2^*$, where $t_i^* = t_i^o/(T_i-w_o)$ for $i=1,2$, such that $t_1^*, t_2^* \in [0,1]$. Once (sub-)class characteristics are removed, the relative optimal locations should be distributed according to a uniform distribution in $[0,1]$. }

@@

More precisely, let us define starting points \(s_i^{(k)}\) for each
signature \(k = 1, 2\) as

\begin{eqnarray}\label{eq.start}
s^{(k)}_i = 
\begin{cases}
t_k^* + i w_v & \text{ for } i < 0 \\
t_k^* + w_ o + i w_v & \text{ for } i \ge 0,
\end{cases}
\end{eqnarray}

for integer values of \(i\) with \(0 < s^{(k)}_i \le T_k - w_v\).

@@

\subsection{A problem with failed
tests}\label{a-problem-with-failed-tests}

Looking closer at \autoref{eq.start}, we see that by definition, some
number of tests will fail to produce a result.
\hh{The problem of failed tests is first mentioned in \citet{afte-chumbley}. Unfortunately, the authors do not provide any percentage of how many tests failed for their data.}
The algorithm fails to produce for two reasons: either the number of
eligible same-shift pairs is zero, or the number of different-shift
pairs is zero.

@@

\subsection{A modified approach}\label{a-modified-approach}

@@

We then define the pairs for different-shifts as

\begin{equation}\label{eq.diff2}
\left(s_{(j)}^{(1)}, s_{(I-j+1)}^{(1)} \right) \text{ for } j = 
\begin{cases}
1, ..., I & \text{ for even } I \\
1, ..., (I-1)/2, (I-1)/2 + 2, ..., I & \text{ for odd } I
\end{cases},
\end{equation}

i.e.~for an odd number of same-shift correlations, we skip the middle
pair for the different-shift correlations (see \autoref{sketch-diff-2}).
This pairing ensures that the number of different-shift pairings is the
same or at most one less than the number of same-shift pairings in all
tests.
\hh{In the remainder of the paper, we will refer to the algorithm defined by \citet{hadler} as {\bf (CS1)} and the suggested modified algorithm as {\bf (CS2)} and compare their performance on the available scans of the Hamby study.}

@@

\section{Testing setup}\label{testing-setup}

\subsection{The Data}\label{the-data}

@@

\subsection{Setup}\label{setup}

@@

\section{Results}\label{results}

\subsection{Failed Tests}\label{failed-tests}

Initially, the default settings suggested in \citet{hadler} were used:
\(w_o = 120\) pixels or about 190 \(\mu m\) (ten percent of the average
length of profiles) and coarseness \(c = 0.25\). \autoref{fig:fails}
shows the percentage of failed tests among the 85,491 land-to-land
comparisons of the NIST data for different values of the validation
window size \(w_v\). For same-source lands up to 12.5 percent of the
tests fail using CS1. The highest percentage of failed tests under CS2
is 1.3\% for different source tests using a validation window size
\(w_v\) of 60 pixels.
\hh{Rates of expected failures are based on simulation runs using covariances between locations of same-source profiles of 0.854, and 0.120 for locations from different-source profiles, matching observed covariances for the Hamby scans. }\hh{ Observed failure rates are higher than expected. This might be due to remaining sub-class structure at a coarseness of 0.3 resulting in a distribution of optimal locations different from the assumed uniform. }

@@

\subsection{Coarseness}\label{coarseness}

@@

\hh{\autoref{fig:profile-sketch} gives an overview of the effect of different coarseness parameters: from left to right, coarseness levels $c$ are varied in steps of 0.05 from 0.1 to 0.3. The top row shows  resulting signature after smoothing the profile shown in Figure~\ref{fig:sigs-profiles} with the coarseness specified. 
The histograms in the bottom row show the relative optimal location $t^*$. Optimal locations are distributed uniformly once (sub-)class characteristics are removed. However, for coarseness values of $c > 0.20$ we see quite distinct boundary effects: optimal locations $t^*$ are found at the very extreme ends of a profile more often than one would expect based on a uniform distribution.}

@@

\subsection{Type II error rates}\label{type-ii-error-rates}

@@

\subsection{Observed versus Nominal Type I error
rates}\label{observed-versus-nominal-type-i-error-rates}

@@

\section{Conclusions}\label{conclusions}

\hh{
We started out trying to assess the suitability of the Chumbley Score's suitability for matching striae on bullet lands. Besides using recommended defaults, we also proposed ways to optimize parameters: based on the assumption that once sub-class characteristics are removed, optimal locations are distributed uniformly across the profile, we found  for bullets the smaller coarseness value of $c = 0.15$ to be suitable. 
Optimized window sizes were found based on cross-validation to minimize type 2 error rates. 
Method CS1 proposed by \citet{hadler} has a minimal type 2 error rate of XXX for an optimized window size of YYY - which is considerably higher than the error rates achieved on matching toolmarks. 
Unfortunately, method CS1 also has a high rate of failed tests -- situations, in which the algorithm does not provide a result, due to the way different-shift pairs are constructed. 
XXX
The algorithm CS2 introduced here is constructed in a way that reduces the number of failures by a magnitude of about XXX. While reducing the failure rate, the algorithm also shows an increase in the power of the test. Type II error rates reach a minimum of XXX for an optimized window size of YYY. While significantly reduced over CS1, CS2 still has type 2 error rates on bullet lands that are  higher than the error rates achieved on the -much larger- toolmarks.
However, bullets usually have multiple lands - in the case of Ruger P85s as used in the Hamby study, there are six lands for each bullet. We might be able to get more power out of the test by adapting CS2 to deal work in a bullet-to-bullet comparison.
}

\hh{
Fix in CS2 should also improve power for matching toolmarks. 
}

>>>>

\newcommand{\hh}[1]{{\textcolor{orange}{#1}}}
\newcommand{\gk}[1]{{\textcolor{blue}{#1}}}
\newcommand{\cited}[1]{{\textcolor{red}{#1}}}
\setlength\parindent{0pt}

@@

\hypertarget{introduction-and-background}{%
\section{Introduction and
Background}\label{introduction-and-background}}

\hypertarget{motivation}{%
\subsection{Motivation}\label{motivation}}

@@

{\textcolor{orange}{In the next page and a half it is easy to lose the red line. It might help to include a table with an overview.
The table should include the reference to the paper, the data used, the statistical method and the associated error rates.}}
{\textcolor{blue}{table included}}

Various toolmarks have been studied in the literature:
\citet{manytoolmarks1} and \citet{chumbley} have been analyzing
screwdriver marks digitized using a profilometer; \citet{manytoolmarks2}
have investigated 3D marks from screwdriver, tongue and groove pliers
captured using a confocal microscope; \citet{afte-chumbley} have been
investigated digitized marks from slip-joint pliers generated by a
surface profilometer.

{\textcolor{orange}{We need an additional sentence here to get from the data to the statistical methods ... }}
{\textcolor{blue}{Analysis of these digitized markings require the use of statistical methods which can quantify the scientific mechanism of comparing markings and serve as metrics for correctly identifying the error rate.}}

\citet{manytoolmarks2} define a relative distance metric and use it as
similarity measure between two toolmarks. \citet{manytoolmarks1} extract
many small segments in the markings of two toolmarks and compare
similiarity using a maximum pearson correlation coefficient. The
Chumbley scoring method, first introduced by \citet{chumbley}, uses a
similar but more extensive framework based on a Mann-Whitney U test of
the resulting correlation coefficients. This approach is
non-deterministic, because segments are chosen randomly. \citep{hadler}
make the score deterministic for each pair of toolmarks by choosing
segments for comparison systematically. This approach also ensures
independence between segements of striae. In this paper, we are
investigating the applicability of the Chumbley scoring method by
\citet{hadler} to assess striation marks on bullet lands for same-source
identification.

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Research paper & Method & Data Source & False Positives & False Negatives\\
  \hline
 \textbf{\citet{manytoolmarks1}} & Maximized & Screwdrivers & & \\ & Correlation &  & - & - \\ \hline
 \textbf{\citet{chumbley}} & Randomized & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 2.3\% & 8.9\%  \\ \hline
  \textbf{\citet{afte-chumbley}} & Randomized & Slip-joint  &  & \\ & Chumbley Score &  & - & - \\ \hline 
  \textbf{\citet{hadler}} & Deterministic & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 0\% & 6\%  \\ \hline
  \textbf{\citet{manytoolmarks2}} & Similarity Measure & Screwdrivers & & \\ (Different Surfaces-same angle) & Relative Distance Metric & & 5.9\% &  9.4\% \\ (Same Surfaces-same angle)&  & & 0.22\% &  0\% \\
   \hline
\end{tabular}}
\caption{\label{tab:toolmarks-ER} Error Rates for Toolmarks using  variations of the chumbley score method}
\end{table}

@@

\hypertarget{scans-for-land-engraved-areas}{%
\subsection{Scans for land engraved
areas}\label{scans-for-land-engraved-areas}}

@@

\hypertarget{the-chumbley-score-test}{%
\subsection{The Chumbley Score Test}\label{the-chumbley-score-test}}

@@

After removing sub-class structure, the Chumbley scores is calculated in
two steps: an optimization step and a validation step. In the
optimization step, the two markings are aligned horizontally such that
within a pre-defined window of length \(w_o\) the correlation between
\(x(t_1)\) and \(y(t_2)\) is maximized: \[
\left(t_1^o, t_2^o\right) = \mathop{\arg \max}\limits_{1 \le t_1 \le T_1, 1 \le t_2 \le T_2} \text{cor} \left(x^{w_o} (t_1), y^{w_o}(t_2) \right)
\] This results in an optimal vertical (in-phase) shift of
\(t_1^o - t_2^o\) for aligning the two markings.
{\textcolor{orange}{We will denote the relative optimal locations as $t_1^*$ and $t_2^*$, where $t_i^* = t_i^o/(T_i-w_o)$ for $i=1,2$, such that $t_1^*, t_2^* \in [0,1]$. Once (sub-)class characteristics are removed, the relative optimal locations should be distributed according to a uniform distribution in $[0,1]$. }}

@@

More precisely, let us define starting points \(s_i^{(k)}\) for each
signature \(k = 1, 2\) as \begin{eqnarray}\label{eq.start}
s^{(k)}_i = 
\begin{cases}
t_k^* + i w_v & \text{ for } i < 0 \\
t_k^* + w_ o + i w_v & \text{ for } i \ge 0,
\end{cases}
\end{eqnarray} for integer values of \(i\) with
\(0 < s^{(k)}_i \le T_k - w_v\).

@@

\hypertarget{a-problem-with-failed-tests}{%
\subsection{A problem with failed
tests}\label{a-problem-with-failed-tests}}

Looking closer at \autoref{eq.start}, we see that by definition, some
number of tests will fail to produce a result.
{\textcolor{orange}{The problem of failed tests is first mentioned in \citet{afte-chumbley}. Unfortunately, the authors do not provide any percentage of how many tests failed for their data.}}
The algorithm fails to produce for two reasons: either the number of
eligible same-shift pairs is zero, or the number of different-shift
pairs is zero.

@@

\hypertarget{a-modified-approach}{%
\subsection{A modified approach}\label{a-modified-approach}}

@@

We then define the pairs for different-shifts as
\begin{equation}\label{eq.diff2}
\left(s_{(j)}^{(1)}, s_{(I-j+1)}^{(1)} \right) \text{ for } j = 
\begin{cases}
1, ..., I & \text{ for even } I \\
1, ..., (I-1)/2, (I-1)/2 + 2, ..., I & \text{ for odd } I
\end{cases},
\end{equation} i.e.~for an odd number of same-shift correlations, we
skip the middle pair for the different-shift correlations (see
\autoref{sketch-diff-2}). This pairing ensures that the number of
different-shift pairings is the same or at most one less than the number
of same-shift pairings in all tests.
{\textcolor{orange}{In the remainder of the paper, we will refer to the algorithm defined by \citet{hadler} as{\bf (CS1)} and the suggested modified algorithm as{\bf (CS2)} and compare their performance on the available scans of the Hamby study.}}

@@

\begin{figure}[hbtp]
\centering
\includegraphics[width=0.62\textwidth]{images/same-shift-failure.png}

\caption{\label{same-shift-failure}Sketch of same-shift pairings  (top) when the lag is too large to accomodate a a vaildation window in either of the two signatures}
\end{figure}

\begin{figure}[hbtp]
\centering
\includegraphics[width=0.7\textwidth]{images/diff-shift-failure.png}

\caption{\label{diff-shift-failure}Sketch of diff-shift pairings  (top) when the number of diff-shift computations is likely to 0}
\end{figure}

\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/CS1-failure-CS2-success.png}
\caption{\label{CS1-failure-CS2-success}Sketch of a case where CS1 fails but CS2 does not fail}
\end{figure}

\hypertarget{testing-setup}{%
\section{Testing setup}\label{testing-setup}}

\hypertarget{the-data}{%
\subsection{The Data}\label{the-data}}

@@

\hypertarget{setup}{%
\subsection{Setup}\label{setup}}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{failed-tests}{%
\subsection{Failed Tests}\label{failed-tests}}

Initially, the default settings suggested in \citet{hadler} were used:
\(w_o = 120\) pixels or about 190 \(\mu m\) (ten percent of the average
length of profiles) and coarseness \(c = 0.25\). \autoref{fig:fails}
shows the percentage of failed tests among the 85,491 land-to-land
comparisons of the NIST data for different values of the validation
window size \(w_v\). For same-source lands up to 12.5 percent of the
tests fail using CS1. The highest percentage of failed tests under CS2
is 1.3\% for different source tests using a validation window size
\(w_v\) of 60 pixels.
{\textcolor{orange}{Rates of expected failures are based on simulation runs using covariances between locations of same-source profiles of 0.854, and 0.120 for locations from different-source profiles, matching observed covariances for the Hamby scans. }}{\textcolor{orange}{ Observed failure rates are higher than expected. This might be due to remaining sub-class structure at a coarseness of 0.3 resulting in a distribution of optimal locations different from the assumed uniform. }}

@@

\hypertarget{coarseness}{%
\subsection{Coarseness}\label{coarseness}}

@@

{\textcolor{orange}{\autoref{fig:profile-sketch} gives an overview of the effect of different coarseness parameters: from left to right, coarseness levels $c$ are varied in steps of 0.05 from 0.1 to 0.3. The top row shows  resulting signature after smoothing the profile shown in Figure~\ref{fig:sigs-profiles} with the coarseness specified. 
The histograms in the bottom row show the relative optimal location $t^*$. Optimal locations are distributed uniformly once (sub-)class characteristics are removed. However, for coarseness values of $c > 0.20$ we see quite distinct boundary effects: optimal locations $t^*$ are found at the very extreme ends of a profile more often than one would expect based on a uniform distribution.}}

{\textcolor{blue}{The key effect of the optimal locations and thereby the coarseness is seen in the number of failed tests. Irrespective of whether CS1 or CS2 is being used, if the relative optimal locations are at the boundaries we will see an increase in the number of failed tests. A balance is therefore needed in the selection of the coarseness parameter which reduces the boundary effect but also does not remove important (sub-)class characteristics}}

@@

\hypertarget{type-ii-error-rates}{%
\subsection{Type II error rates}\label{type-ii-error-rates}}

@@

\hypertarget{observed-versus-nominal-type-i-error-rates}{%
\subsection{Observed versus Nominal Type I error
rates}\label{observed-versus-nominal-type-i-error-rates}}

@@

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

{\textcolor{orange}{
We started out trying to assess the suitability of the Chumbley Score's suitability for matching striae on bullet lands. Besides using recommended defaults, we also proposed ways to optimize parameters: based on the assumption that once sub-class characteristics are removed, optimal locations are distributed uniformly across the profile, we found  for bullets the smaller coarseness value of $c = 0.15$ to be suitable. 
Optimized window sizes were found based on cross-validation to minimize type 2 error rates. 
Method CS1 proposed by \citet{hadler} has a minimal type 2 error rate of XXX for an optimized window size of YYY - which is considerably higher than the error rates achieved on matching toolmarks. 
Unfortunately, method CS1 also has a high rate of failed tests -- situations, in which the algorithm does not provide a result, due to the way different-shift pairs are constructed. 
XXX
The algorithm CS2 introduced here is constructed in a way that reduces the number of failures by a magnitude of about XXX. While reducing the failure rate, the algorithm also shows an increase in the power of the test. Type II error rates reach a minimum of XXX for an optimized window size of YYY. While significantly reduced over CS1, CS2 still has type 2 error rates on bullet lands that are  higher than the error rates achieved on the -much larger- toolmarks.
However, bullets usually have multiple lands - in the case of Ruger P85s as used in the Hamby study, there are six lands for each bullet. We might be able to get more power out of the test by adapting CS2 to deal work in a bullet-to-bullet comparison.
}}

{\textcolor{orange}{
Fix in CS2 should also improve power for matching toolmarks. 
}}
